Plasticity and adaptation to high light intensity amplify the advantage of amphistomatous leaves

### 2025-08-19 ms/ms.tex
migrating main text to word

\begin{center}
$^1$School of Life Sciences, University of Hawaiʻi, Mānoa, Hawaiʻi, USA\\
$^2$Department of Botany, University of Wisconsin, Madison, Wisconsin, USA\\
\medskip
*Corresponding author: cdmuir@wisc.edu\\
\medskip
ORCID: Christopher D. Muir — \href{https://orcid.org/0000-0003-2555-3878}{0000-0003-2555-3878}\\
\end{center}

@@

The presence of stomata on both leaf surfaces (amphistomy) can increase
photosynthesis in C\(_3\) plants by reducing the path length for
CO\(_2\) diffusion between substomatal cavities and chloroplasts.
Paradoxically, despite this photosynthetic advantage, most leaves are
hypostomatous (stomata on lower surface only). Based on the observation
that amphistomatous leaves are most common among herbaceous plants
growing in sunny habitats, including many crop species, it has long been
hypothesized that traits affecting CO\(_2\) supply or demand amplify the
advantage of amphistomy under high light intensity. We used a recently
developed method to quantify amphistomy advantage, the photosynthetic
rate of an amphistomatous leaf relative to an otherwise identical
hypostomatous leaf, in 29 diverse populations representing 14 species of
wild tomatoes (\emph{Solanum} sect. \emph{Lycopersicon} and sect.
\emph{Lycopersicoides}). Plants grown under high light intensity benefit
more from amphistomy in terms of CO\(_2\) assimilated for a given
stomatal conductance than plants grown under low light, regardless of
measurement light intensity. Furthermore, populations native to open
habitats tend to benefit more from amphistomy compared to those from
more closed habitats. Developmental plasticity and adaptation of leaves
to high light intensity likely play a combined role in explaining the
adaptive significance and distribution of amphistomy in nature. We
estimate that hypostomatous leaves would expend 10-65\% more water to
achieve the same photosynthetic rate as a comparable amphistomatous
leaf, implying large fitness costs of amphistomy in ecological contexts
where this trait is rare.

@@

Stomata are microscopic pores on the surfaces of leaves and other
photosynthetic organs formed by a pair of guard cells. They are
essential for balancing carbon gained per unit water lost and were an
essential innovation in vascular plants, permitting them to grow tall on
land by enabling access to CO\(_2\) for photosynthetic carbon
assimilation while preventing hydraulic failure in variable environments
(\emph{1}--\emph{3}). Optimal stomatal function depends on both dynamic
changes in aperture on the scale of minutes to hours, as well as static
anatomy determined by developmental plasticity and constitutive genetic
differences (\emph{4}--\emph{6}). Understanding how stomata respond to
environmental change over daily, developmental, and evolutionary time is
important for studying adaptation (\emph{1}, \emph{7}, \emph{8}),
inferring paleoclimate from fossil cuticles (\emph{9}--\emph{11}),
predicting responses to climate change (\emph{12}--\emph{15}), and
improving crops (\emph{16}). Stomatal function contributes to global
carbon and water cycles (\emph{17}) and therefore predicting future
climate (\emph{18}).

@@

\emph{Acclimation hypothesis}: Photosynthetic induction to high light
intensity typically involves increases in total leaf stomatal
conductance (increased CO\(_2\) supply), the concentration of active
Rubisco, and electron transport capacity (increased CO\(_2\) demand). If
the acclimation hypothesis is correct, we predict that
\(\aahigh{} > \aalow{}\) for all populations regardless of native
habitat or growth environment.

\emph{Plasticity hypothesis}: Individuals of the same genotype often
develop dramatically different leaves in sun and shade conditions
(\emph{47}). Plastic responses are likely adaptations to optimize
photosynthesis at different light intensities in variable environments
(\emph{48}). Plastic changes in leaf anatomy and biochemistry could
modulate \aax{} as a byproduct. Thicker or less porous leaves, both of
which are associated with high leaf mass per area (LMA), will have lower
\gias; leaves with increased total stomatal density and photosynthetic
capacity have greater potential CO\(_2\) supply and demand. Under the
plasticity hypothesis, we predict that \(\aasun{} > \aashade{}\) for all
populations and light intensities.

@@

\caption{\label{fig-hypotheses}\textbf{Conceptual outline of three nonmutually exclusive hypotheses explaining why amphistomy advantage (\aax{}) might be greater for leaves in sunny, open habitats.}
The acclimatopn hypothesis predicts that \aax{} is greater under high
measurement light intensity (\ppfdqty{2000}) than low measurement light
intensity (\ppfdequals{150}), regardless of growth light intensity or
native plant area index (PAI). The plasticity hypothesis predicts that
\aax{} is greater for plants grown in sun than shade, regardless of
measurement light intensity or native PAI. The constitutive hypothesis
predicts that \aax{} is greater for plants adapted to sunny habitats
(lower native PAI) than shaded habitats, regardless of measurement light
intensity or growth light intensity.}

@@

We tested these hypotheses by comparing \aax{} among amphistomatous wild
tomato species (Table~\ref{tbl-populations}; Figure~\ref{fig-sr-rxnorm};
(\emph{50})) from different native light habitats, grown under simulated
sun and shade light treatments, and measured under contrasting light
intensities (low and high). We measured \aax{} on 572 individual plants
from 29 populations (average of 9.86 replicates per light treatment)
using a recently developed method (\emph{46}). With this method, we
directly compare the photosynthetic rate of an untreated amphistomatous
leaf to that of the same leaf with gas exchange blocked through the
adaxial (upper) surface by transparent plastic, which we refer to as
`pseudohypostomy'. To compare amphi- and pseudohypostomatous leaves at
identical whole-leaf stomatal conductance (\gsw), we measure \(A\) over
a range of \gsw, inducing stomatal opening and closure by modulating
humidity (see Materials and Methods for further details). We estimated
`amphistomy advantage' (\aax) \emph{sensu} (\emph{20}), but with
modifications previously described in (\emph{46}) and here (Materials
and Methods). The native light intensity was represented by plant area
index (PAI \(\unit{\meter\squared\per\meter\squared}\)), estimated using
a global gridded data set derived from the Global Ecosystem Dynamics
Investigation {[}GEDI; (\emph{51}){]} and georeferenced accession
collection information from the Tomato Genetics Resource Center
(Table~\ref{tbl-populations}). The growth light intensities were
\ppfdequals{761} (sun treatment) and \ppfdqty{115} (shade treatment)
while all other environment conditions were nearly identical (see
Materials and Methods). The high and low measurement light intensities
were \ppfdequals{2000} (97.8:2.24 red:blue) and \ppfdequals{150}
(87.0:13.0 red:blue), respectively.

@@

The pattern of \aax{} across wild tomatoes strongly supports the
plasticity hypothesis, contradicts the acclimation hypothesis, and
provides modest support for the constitutive hypothesis. Plastic changes
in leaf thickness and/or packing density, summarized by the bulk leaf
mass per area (LMA), may mediate the effect of growth light intensity on
\aax{}. LMA increased in sun grown plants in all populations by an
average of 123\% {[}95\% CI: 42.9 to 256\%{]}, qualitatively similar to
plastic responses in many species (\emph{47}). While LMA is weakly,
albeit significantly (Table~\ref{tbl-fit_aa2}), associated with
individual-level \aax{} (Fig.~\ref{fig-lma_aa}), the effect of growth
light intensity on \aax{} is still predictive based on model comparison
(Table~\ref{tbl-aa_loo1}). However, the direct effect of the sun
treatment on \aax{} was weaker when LMA was included in the model
(Table~\ref{tbl-fit_aa1} vs. Table~\ref{tbl-fit_aa2}), suggestive of a
mediating role. Many anatomical traits underlie LMA (\emph{52}) and
future research will be needed to identify which particular traits, such
as leaf thickness or mesophyll porosity, are responsible for mediating
\aax{}. The fact that the \aax{} of sun plants was greater under low
measurement light intensity supports a long-standing hypothesis that
resistance to CO\(_2\) diffusion is greater in the upper than lower
portions of the leaf interior (\emph{53}). At low light intensity,
photosynthesis is weighted toward the upper palisade where most light is
intercepted. If resistance to CO\(_2\) diffusion is high in the upper
palisade, amphistomy may, unexpectedly, be particularly beneficial for
sun leaves experiencing intermittent shade or cloud cover. Our study is
limited in testing this because we could not directly measure the
stomatal conductance ratio and intercellular resistance on each surface.
Future experiments measuring \aax{} with a dual sided chamber
(\emph{31}, \emph{33}) can overcome these limitations.

@@

\caption{\label{fig-lma_aa}\textbf{Plastic changes in leaf mass per area (LMA) mediate the effect of sun- to shade-grown plants on amphistomy advantage (\aax{}).}
Each point is the estimated population-level LMA (\(x\)-axis) and \aax{}
(\(y\)-axis) for each population grown under high light intensity
(\ppfdequals{2000}) and low light intensity (\ppfdequals{150}). The
points are colored by growth light treatment. The solid line is the
posterior median of the linear regression of individual-level LMA on
\(\aahigh\) with the shaded region showing the 95\% confidence ribbon;
the dashed line is the same for \(\aalow\). Confidence intervals for
each population estimate were omitted for visual clarity.}

>>>>

\begin{center}
$^1$School of Life Sciences, University of Hawaiʻi; Mānoa, Hawaiʻi, USA\\
$^2$Department of Botany, University of Wisconsin, Madison; Wisconsin, USA\\
\medskip
*Corresponding author: cdmuir@wisc.edu\\
\medskip
ORCID: Christopher D. Muir — \href{https://orcid.org/0000-0003-2555-3878}{0000-0003-2555-3878}\\
\end{center}

@@

The presence of stomata on both leaf surfaces (amphistomy) increases
photosynthesis by reducing the distance for CO\(_2\) diffusion between
stomata and chloroplasts. Paradoxically, most leaves are hypostomatous
(stomata on lower surface), despite the photosynthetic advantage of
amphistomy. Across 29 diverse populations of ``wild tomatoes'', leaves
developed under high light intensity benefit more from amphistomy in
terms of CO\(_2\) assimilated for a given stomatal conductance than
plants developed under low light. Furthermore, populations native to
open habitats benefit more from amphistomy than those from more closed
habitats. Thus, plasticity and adaptation together may explain why
amphistomatous leaves are prevalent in sunny, open habitats, including
many crops. Contrary to common assumptions, amphistomy can save water
because hypostomatous leaves evaporate 10-65\% more to achieve the same
photosynthetic rate.

@@

Stomata are microscopic pores on the surfaces of leaves and other
photosynthetic organs formed by a pair of guard cells. They are
essential for balancing carbon gained per unit water lost and were an
essential innovation in vascular plants, permitting them to grow tall on
land by enabling access to CO\(_2\) for photosynthetic carbon
assimilation while preventing hydraulic failure in variable environments
(\emph{1}--\emph{3}). Optimal stomatal function depends on both dynamic
changes in aperture on the scale of minutes to hours, as well as static
anatomy determined by developmental plasticity and constitutive genetic
differences (\emph{4}--\emph{6}). Understanding how stomata respond to
environmental change over daily, developmental, and evolutionary time is
important for studying adaptation (\emph{1}, \emph{7}, \emph{8}),
inferring paleoclimate from fossil cuticles (\emph{9}--\emph{11}),
predicting responses to climate change (\emph{12}--\emph{15}), and
improving crops (\emph{16}). Stomatal function contributes to global
carbon and water cycles (\emph{17}) and therefore to predicting future
climate (\emph{18}).

@@

\emph{Acclimation hypothesis}: Photosynthetic induction to high light
intensity typically involves increases in total leaf stomatal
conductance (increased CO\(_2\) supply), the concentration of active
Rubisco, and electron transport capacity (increased CO\(_2\) demand). If
the acclimation hypothesis is correct, we predict that
\(\aahigh{} > \aalow{}\) for all populations regardless of native
habitat or growth environment. \(\aahigh{}\) is the \aax{} measured
under high light intensity (\ppfdequals{2000}); \(\aalow{}\) is the
\aax{} measured under low light intensity (\ppfdequals{150}).

\emph{Plasticity hypothesis}: Individuals of the same genotype often
develop dramatically different leaves in sun and shade conditions
(\emph{47}). Plastic responses are likely adaptations to optimize
photosynthesis at different light intensities in variable environments
(\emph{48}). Plastic changes in leaf anatomy and biochemistry could
modulate \aax{} as a byproduct. Thicker or less porous leaves, both of
which are associated with high leaf mass per area (LMA), will have lower
\gias; leaves with increased total stomatal density and photosynthetic
capacity have greater potential CO\(_2\) supply and demand. Under the
plasticity hypothesis, we predict that \(\aasun{} > \aashade{}\) for all
populations and light intensities. \(\aasun{}\) is the \aax{} measured
on sun leaves; \(\aashade{}\) is the \aax{} measured on shade leaves.

@@

\caption{\label{fig-hypotheses}\textbf{Conceptual outline of three nonmutually exclusive hypotheses explaining why amphistomy advantage (\aax{}) might be greater for leaves in sunny, open habitats.}
The acclimation hypothesis predicts that \aax{} is greater under high
measurement light intensity (\ppfdequals{2000}) than low measurement
light intensity (\ppfdequals{150}), regardless of growth light intensity
or native plant area index (PAI). The plasticity hypothesis predicts
that \aax{} is greater for plants grown in sun than shade, regardless of
measurement light intensity or native PAI. The constitutive hypothesis
predicts that \aax{} is greater for plants adapted to sunny habitats
(lower native PAI) than shaded habitats, regardless of measurement light
intensity or growth light intensity.}

@@

We tested these hypotheses by comparing \aax{} among amphistomatous wild
tomato species (Table~\ref{tbl-populations}; Figure~\ref{fig-sr-rxnorm};
(\emph{50})) from different native light habitats, grown under simulated
sun and shade light treatments, and measured under contrasting light
intensities (low and high). We measured \aax{} on 572 individual plants
from 29 populations (average of 9.86 replicates per light treatment)
using a recently developed method (\emph{46}). With this method, we
directly compare the photosynthetic rate of an untreated amphistomatous
leaf to that of the same leaf with gas exchange blocked through the
adaxial (upper) surface by transparent plastic, which we refer to as
`pseudohypostomy'. To compare amphi- and pseudohypostomatous leaves at
identical whole-leaf stomatal conductance (\gsw), we measure \(A\) over
a range of \gsw, inducing stomatal opening and closure by modulating
humidity (see supplementary text for further details). We estimated
`amphistomy advantage' (\aax) \emph{sensu} (\emph{20}), but with
modifications previously described in (\emph{46}) and here
(Supplementary Materials). The native light intensity was represented by
plant area index (PAI \(\unit{\meter\squared\per\meter\squared}\)),
estimated using a global gridded data set derived from the Global
Ecosystem Dynamics Investigation {[}GEDI; (\emph{51}){]} and
georeferenced accession collection information from the Tomato Genetics
Resource Center (Table~\ref{tbl-populations}). The growth light
intensities were \ppfdequals{761} (sun treatment) and \ppfdqty{115}
(shade treatment) while all other environment conditions were nearly
identical (see Materials and Methods). The high and low measurement
light intensities were \ppfdequals{2000} (97.8:2.24 red:blue) and
\ppfdequals{150} (87.0:13.0 red:blue), respectively.

@@

The pattern of \aax{} across wild tomatoes strongly supports the
plasticity hypothesis, contradicts the acclimation hypothesis, and
provides modest support for the constitutive hypothesis. Plastic changes
in leaf thickness and/or packing density, summarized by the bulk leaf
mass per area (LMA), may mediate the effect of growth light intensity on
\aax{}. LMA increased in sun grown plants in all populations by an
average of 123\% {[}95\% CI: 42.9 to 256\%{]}, qualitatively similar to
plastic responses in many species (\emph{47}). While LMA is weakly,
albeit significantly (Table~\ref{tbl-fit_aa2}), associated with
individual-level \aax{} (Fig.~\ref{fig-lma_aa}), the effect of growth
light intensity on \aax{} is still predictive based on model comparison
using information criteria (Table~\ref{tbl-aa_loo1}). However, the
direct effect of the sun treatment on \aax{} was weaker when LMA was
included in the model (Table~\ref{tbl-fit_aa1} vs.
Table~\ref{tbl-fit_aa2}), suggestive of a mediating role. Many
anatomical traits underlie LMA (\emph{52}) and future research will be
needed to identify which particular traits, such as leaf thickness or
mesophyll porosity, are responsible for mediating \aax{}. The fact that
the \aax{} of sun plants was greater under low measurement light
intensity supports a long-standing hypothesis that resistance to
CO\(_2\) diffusion is greater in the upper than lower portions of the
leaf interior (\emph{53}). At low light intensity, photosynthesis is
weighted toward the upper palisade where most light is intercepted. If
resistance to CO\(_2\) diffusion is high in the upper palisade,
amphistomy may, unexpectedly, be particularly beneficial for sun leaves
experiencing intermittent shade or cloud cover. Our study is limited in
testing this because we could not directly measure the stomatal
conductance ratio and intercellular resistance on each surface. Future
experiments measuring \aax{} with a dual sided chamber (\emph{31},
\emph{33}) can overcome these limitations.

@@

\caption{\label{fig-lma_aa}\textbf{Plastic changes in leaf mass per area (LMA) partially mediate the effect of sun- to shade-grown plants on amphistomy advantage (\aax{}).}
Each point is the estimated LMA (\(x\)-axis) and \aax{} (\(y\)-axis) for
each population grown under high light intensity (\ppfdequals{2000}) and
low light intensity (\ppfdequals{150}). The points are colored by growth
light treatment. The solid line is the posterior median of the linear
regression of individual-level LMA on \(\aahigh\) with the shaded region
showing the 95\% confidence ribbon; the dashed line is the same for
\(\aalow\). Confidence intervals for each population estimate were
omitted for visual clarity.}
